Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study

### 2015-03-07 chemotaxis/paper/draft1.tex
updated stuff after ronen visited

\journal{Computers \& Chemical Engineering}

@@

\begin{itemize}

\item
Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc).

\item
Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.

\item
Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").

\item
Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome").

\item
However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.

\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density

\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections.

\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example. The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g. in the chemotaxis, it will not necessarily be the two fluxes).

\item
In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.

\item
All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline. We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).

\end{itemize}

@@

@@

We first consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes,  the principal axes parameterize the manifold.
%
However, in the case when the manifold is curved, it is more difficult to define a set of coordinates.

In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator can be used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
%
To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the rectangular domain \cite{singer2008non}. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}). 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
%
Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions (TODO: is it explicit in the construction?). 
%
The eigenvectors of this matrix then approximate the eigenfunctions of this operator \cite{...}.

Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathcal{M}_d$. 
%
We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
$\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator (with Neumann boundary conditions) on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}
%
As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.

@@

@@

@@

In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the leading eigenvectors provide us with a parameterization of the underlying manifold.
%
However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
%
Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_1  > L_2$. 
%
We recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left( \frac {\pi z_1}{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left( \frac {\pi z_2}{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}). 
%
In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a repeated eigendirection of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Utilizing only the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics yields the most parsimonious representation of the data.
%
We will show how we can automatically detect these modes to obtain a meaningful representation of the data.
%
Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections. 
%
This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first principal direction at each iteration. 
%
However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
%
Here, we propose an alternative approach to address the problem of repeated eigendirections. 
%
We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.

@@

\eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
%
We begin by computing the matrix
\begin{equation}
X_{k-1}(i) = \begin{bmatrix}
1 & \phi_1(1) - \phi_1(i) & \dots & \phi_{k-1}(1)- \phi_{k-1}(i) \\
1 & \phi_1(2) - \phi_1(i) & \dots & \phi_{k-1}(2)- \phi_{k-1}(i) \\
\vdots & \vdots & \ddots & \vdots \\
1 & \phi_1(m) - \phi_1(i) & \dots & \phi_{k-1}(m)- \phi_{k-1}(i) 
\end{bmatrix}
\end{equation}
%
and the matrix 
\begin{equation}
W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right).
\end{equation}
%
We then calculate the smoothing matrix $L_{k-1} \in \mathbb{R}^{m \times m}$, where 
\begin{equation}
L_{k-1} =
\begin{bmatrix}
e_1^T \left( X_{k-1}^T(1) W_{k-1}(1) X_{k-1}(1) \right) ^{-1} X_{k-1}^T(1) W_{k-1}(1) \\
e_1^T \left( X_{k-1}^T(2) W_{k-1}(2) X_{k-1}(2) \right) ^{-1} X_{k-1}^T(2) W_{k-1}(2) \\
\vdots \\
e_1^T \left( X_{k-1}^T(m) W_{k-1}(m) X_{k-1}(m) \right) ^{-1} X_{k-1}^T(m) W_{k-1}(m) \\
\end{bmatrix},
\end{equation}
%
where $e_1^T = (1, 0, \dots, 0)$. 
%
We can then write
%
\begin{equation} 
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
\end{equation}

@@

Once we have determined which eigenvectors parameterize independent directions along the manifold, we can use the corresponding eigenvalues to calculate the relative lengths along these directions. 
%
Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics (i.e., those eigenvectors where $r_{i_j}$ is large). 
%
From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths along the manifold by
\begin{equation} \label{eq:est_lengths}
L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}
\end{equation}
where $L_1, L_2, \dots$ are the relative lengths along the principal directions of the manifold. 
%
We can use this measure to determine how many components are required to retain most of the information within the data set: if $L_j$ is small, then this component can be disregarded (TODO: define small).

@@

We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on three synthetic data sets. 
%
The first data set is a two-dimensional strip where the eigenvalues and eigenvectors are known analytically.
%
The second and third data sets are nonlinear manifolds which demonstrate the flexibility of our approach.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L2}
\includegraphics[height=3cm]{strip_spectrum_L2}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L4}
\includegraphics[height=3cm]{strip_spectrum_L4}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=2.5cm]{strip_data_L8}
\includegraphics[height=3cm]{strip_spectrum_L8}
\caption{}
\end{subfigure}
%
\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$. The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red. From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.  }
\label{fig:strip_compare_analytic}
\end{figure}

>>>>

%\journal{Computers \& Chemical Engineering}
\journal{Applied and Computational Harmonic Analysis Letter}

@@

In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{....}. 
%
In particular, for dynamical systems, data-driven methodologies have been shown to be essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
%
For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%
For such cases, data obtained from observations and/or simulations of the dynamical system combined with these methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.

%For many engineering applications, it is often advantageous or even essential to effectively parameterize a system's dynamics. 
%TODO: add reasons why, potential applications, etc.
%%
%Often, such a parameterization cannot be obtained analytically, due to the complexity of the system.
%%
%For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
%%
%In such cases, we utilize data obtained from observations and/or simulations of the dynamical system, and turn to data-driven analysis methods to provide an accurate low-dimensional description of this data.
%%
%This low-dimensional description can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior. 
%

Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven analysis technique, to analyze data.
%
Most manifold learning algorithms \cite{...} construct parameterizations of the data through the spectral analysis of a Laplace operator.
%
The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
%
Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale ab-initio model.

These algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
%
Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics. 
%
From a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are 	``repeated eigendirections'' which describe the same mode. 
%
Identifying those eigenvectors is critical for obtaining an accurate parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
%
This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections.

In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression.
%
We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator.  
%
This will allow us to validate our proposed approach.
%
We then consider a complex data set from a stochastic dynamical system which models cellular chemotaxis. 
%
The recent advances in observers and metrics, coupled with the proposed approach for factoring out repeated eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
%
This pipeline will allow us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.

%\begin{itemize}
%
%\item
%Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc). 
%
%\item
%Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
%
%\item
%Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
%
%\item
%Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome"). 
%
%\item
%However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
%
%\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
%
%\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections. 
%
%\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example. The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g. in the chemotaxis, it will not necessarily be the two fluxes).  
%
%\item
%In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
%
%\item
%All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline. We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).
%
%\end{itemize}

@@

Data from dynamical systems is most often high-dimensional, since it is collected at the microscale.
%
However, the macroscale dynamics often evolve to a low-dimensional description. 
%
In such cases, the high-dimensional microscale data will be restricted to a low-dimensional manifold embedded in the high-dimensional space. 
%
Extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics \cite{...}.

Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
%
We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$. 
%
We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.

@@

We start by considering a continuous setting where rigorous analysis is available for specific examples. 
%
In such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial. 
%
The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis. 
%
Thus, we consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
%
For linear hyperplanes, the principal axes parameterize the manifold.
%
However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent.

In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
%
To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are 
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}. 
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}). 
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper. 
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.

@@

In most applications, we are not given a description of the continuous manifold. 
%
Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
%
We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
%
It was shown in \cite{...} that, in the limit of infinite data, the discrete Laplacian constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator \cite{...}.

Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix $W \in \mathbb{R}^{m \times m}$, with
\begin{equation} \label{eq:W}
W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
\end{equation}
where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
%
The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
%
We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
%
In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}. 
%
Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
\begin{equation} \label{eq:evals_relationship}
\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
\end{equation}
%

@@

As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{...}.
%
We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.%, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
%
Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.

@@

From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.  
%
From \eqref{eq:evals}, we see that if we sort the eigenvectors by the magnitude of the corresponding eigenvalues, these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections. 
%
However, these two eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
%
In practice, identifying which eigenvectors correspond to unique eigendirections is nontrivial, as the correspondence between the original data and the coordinates which parameterize the manifold are unknown.
%
We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold cannot easily be obtained from the coordinates of the original data.

@@

Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues, assuming that the leading eigenvectors provide a parameterization of the underlying manifold.
%
However, as discussed in the previous section, some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
%
For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$,
recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$. 
%
However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete operator via \eqref{eq:evals_relationship}). 
%
In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a repeated eigendirection of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
%
We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
%
Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields the most parsimonious representation of the data.
%
We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data.
%
Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along these unique eigendirections.

@@

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections. 
%
This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
%
However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
TODO: find other advantages of our proposed method. 
%
Here, we propose an alternative approach to address the problem of repeated eigendirections. 
%
We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
%
If the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
%
We will use a local linear function 
\begin{equation}
\phi_k \approx \alpha + \beta^T \Phi_{k-1}
\end{equation}
%
as our functional approximation, where 
%
$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$. 
%
The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.

@@

The error in \eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
%
We begin by computing the matrix
\begin{equation}
X_{k-1}(i) = \begin{bmatrix}
1 & \phi_1(1) - \phi_1(i) & \dots & \phi_{k-1}(1)- \phi_{k-1}(i) \\
1 & \phi_1(2) - \phi_1(i) & \dots & \phi_{k-1}(2)- \phi_{k-1}(i) \\
\vdots & \vdots & \ddots & \vdots \\
1 & \phi_1(m) - \phi_1(i) & \dots & \phi_{k-1}(m)- \phi_{k-1}(i) 
\end{bmatrix}
\end{equation}
%
and the matrix 
\begin{equation}
W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right).
\end{equation}
%
We then calculate the smoothing matrix $L_{k-1} \in \mathbb{R}^{m \times m}$, where 
\begin{equation}
L_{k-1} =
\begin{bmatrix}
e_1^T \left( X_{k-1}^T(1) W_{k-1}(1) X_{k-1}(1) \right) ^{-1} X_{k-1}^T(1) W_{k-1}(1) \\
e_1^T \left( X_{k-1}^T(2) W_{k-1}(2) X_{k-1}(2) \right) ^{-1} X_{k-1}^T(2) W_{k-1}(2) \\
\vdots \\
e_1^T \left( X_{k-1}^T(m) W_{k-1}(m) X_{k-1}(m) \right) ^{-1} X_{k-1}^T(m) W_{k-1}(m) \\
\end{bmatrix},
\end{equation}
%
where $e_1^T = (1, 0, \dots, 0)$. 
%
We can then write
%
\begin{equation} 
r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
\end{equation}

@@

Once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions. 
%
Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to eigenvectors which parameterize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large). 
%
From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
\begin{equation} \label{eq:est_lengths}
L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}.
\end{equation}
%
This measure is used to determine how many components are required to retain most of the information within the data set.
%
For example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.

@@

We will demonstrate our proposed approach on three synthetic data sets. 
%
The first data set is the two-dimensional strip discussed previously, where the eigenvalues and eigenvectors are known analytically.
%
The second and third data sets are nonlinear manifolds which demonstrate the flexibility of our approach.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L2}
\includegraphics[height=3cm]{strip_spectrum_L2}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.3\textwidth}
\includegraphics[height=2.5cm]{strip_data_L4}
\includegraphics[height=3cm]{strip_spectrum_L4}
\caption{}
\end{subfigure}
%
\hfill
%
\begin{subfigure}{0.35\textwidth}
\includegraphics[height=2.5cm]{strip_data_L8}
\includegraphics[height=3cm]{strip_spectrum_L8}
\caption{}
\end{subfigure}
%
\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$. The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red. From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.  TODO: normalize colors}
\label{fig:strip_compare_analytic}
\end{figure}

Our first illustrative example consists of three different two-dimensional strip data sets. 
%
Each data set contains $m=1500$ data points uniformly sampled from the strip. 
%
Figure~\ref{fig:strip_compare_analytic} shows the data sets and corresponding eigenspectra.
%
The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a repeated eigendirection, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data. 
%
The analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}) are also indicated in red.
